---
title: "Google Data Analytics Capstone Project."
author: "Shruti Ghiya"
date: "2022-12-21"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

# Introduction

I have recently completed the Google Data Analytics Certification
Program on Coursera. The Final module of the program is a capstone
project which is a showcase of my learning so far. The tools I chose to
use in this project is R programming.

## Scenario You are a junior data analyst working in the marketing

Analyst team at Cyclistic, a bike-share company in Chicago. Until now,
Cyclistic's marketing strategy relied on building general awareness and
appealing to broad consumer segments. One approach that helped make
these things possible was the flexibility of its pricing plans:
single-ride passes, full-day passes, and annual memberships. Customers
who purchase single-ride or full-day passes are referred to as casual
riders. Customers who purchase annual memberships are Cyclistic members.
Cyclistic's finance analysts have concluded that annual members are much
more profitable than casual riders. Although the pricing flexibility
helps Cyclistic attract more customers, The director of marketing
believes that maximizing the number of annual members will be key to
future growth. Rather than creating a marketing campaign that targets
all-new customers, she believes there is a very good chance to convert
casual riders into members. She notes that casual riders are already
aware of the Cyclistic program and have chosen Cyclistic for their
mobility needs.

## Objective:

-   How do annual members and casual riders use Cyclistic bikes
    differently?
-   Why would casual riders buy Cyclistic annual memberships?
-   How can Cyclistic use digital media to influence casual riders to
    become members?

## Prepare

### About Dataset

This is public data that was use to explore how different customer types
are using Cyclistic bikes. This data has been made available by Motivate
International Inc. Dataset downloaded from the below link:
<https://divvy-tripdata.s3.amazonaws.com/index.htm> For the analysis, I
used just Divvy_trips data for quarter Q2 2019 -- Q2 2020. The datasets
have a different name because Cyclistic is a fictional company. The data
is reliable because it was directly downloaded from AWS server and it is
comprehensive current and cited. Data has some limitations and privacy
issues that prohibit from using riders' personally identifiable
information. This means that we won't be able to connect pass purchases
to credit card numbers to determine if casual riders live in the
Cyclistic service area or if they have purchased multiple single passes.

### Files Used :

-   Divvy_Trips_2019_Q2
-   Divvy_Trips_2019_Q3
-   Divvy_Trips_2019_Q4
-   Divvy_Trips_2020_Q1

File format .csv and has the following column names:

-   ride_id • started_at • ended_at • rideable_type • duration •
    start_station_id • start_station_name • end_station_id •
    end_station_name • member_causal • gender • birthyear

## Process

Tools used : RStudio is used for data cleaning and analyzing. Markdown
report is created to verify data is clean and ready to analyze.

### Collect Data

\*\* Installing Package \*\*

```{r}
install.packages("tidyverse", repo="http://cran.rstudio.com/")
install.packages("lubridate", repo="http://cran.rstudio.com/")
library(tidyverse)
library(lubridate)
```

\*\* Loading Data \*\*

```{r}

q2_2019 <- read.csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read.csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read.csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read.csv("Divvy_Trips_2020_Q1.csv")
```

### Warangling Data and Combining it into single file

Checking the column names for each data set.

```{r}
colnames(q2_2019)
colnames(q3_2019)
colnames(q4_2019)
colnames(q1_2020)
```

While checking the structure and column names have noticed that the
column names are not consistent in the data. And to join the data,
columns should match. So renaming the columns matching q1_2020 below:

```{r}
q4_2019 <- rename(q4_2019,
                  ride_id= trip_id,
                  rideable_type = bikeid,
                  started_at = start_time,
                  ended_at = end_time,
                  start_station_name = from_station_name,
                  start_station_id = from_station_id,
                  end_station_name = to_station_name,
                  end_station_id = to_station_id,
                  member_casual = usertype)
```

```{r}
q3_2019 <- rename(q3_2019,
                  ride_id= trip_id,
                  rideable_type = bikeid,
                  started_at = start_time,
                  ended_at = end_time,
                  start_station_name = from_station_name,
                  start_station_id = from_station_id,
                  end_station_name = to_station_name,
                  end_station_id = to_station_id,
                  member_casual = usertype)
```

```{r}

q2_2019 <- rename(q2_2019
                  ,ride_id = "X01...Rental.Details.Rental.ID"
                  ,rideable_type = "X01...Rental.Details.Bike.ID"  
                  ,started_at = "X01...Rental.Details.Local.Start.Time"
                  ,ended_at = "X01...Rental.Details.Local.End.Time"  
                  ,start_station_name = "X03...Rental.Start.Station.Name" 
                  ,start_station_id = "X03...Rental.Start.Station.ID"
                  ,end_station_name = "X02...Rental.End.Station.Name" 
                  ,end_station_id = "X02...Rental.End.Station.ID"  
                  ,member_casual = "User.Type")
```

Inspecting the data frame and looking for incongruities

```{r}
str(q2_2019)
str(q3_2019)
str(q4_2019)
str(q1_2020)
```

ride_id and ride-able_type have inconsistent data type within the
quarterly data. Converting them to character so that can be stacked
properly.

```{r}
q4_2019 <- mutate(q4_2019, ride_id = as.character(ride_id),
                  rideable_type = as.character(rideable_type)) 

q3_2019 <- mutate(q3_2019, ride_id = as.character(ride_id),
                  rideable_type = as.character(rideable_type)) 

q2_2019 <- mutate(q2_2019, ride_id = as.character(ride_id),
                  rideable_type = as.character(rideable_type)) 

```

Joining the individual quarter data into 1 single big data frame.

```{r}
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)
```

There were few columns which were dropped beginning in 2020: \* lat \*
long \* birthyear \* gender

Removing those columns for consistency.

```{r}
all_trips <- all_trips %>% 
  select(-c(start_lat, start_lng, end_lat, end_lng,birthyear, gender,
            "X01...Rental.Details.Duration.In.Seconds.Uncapped",
            "X05...Member.Details.Member.Birthday.Year","Member.Gender",
            tripduration))
```

### Cleaning and adding data to prepare for analysis

Inspecting the new table that has been created

```{r}
nrow(all_trips)
ncol(all_trips)
dim(all_trips)
head(all_trips)
tail(all_trips)
str(all_trips)
summary(all_trips)
```

While checking data, have noticed few problems which we need to fix:

-   The member_casual column: There are different names for members ("
    Subscriber", "member") and for causal riders ("Customer", "casual")

```{r}
unique(all_trips$member_casual)
```

Consolidating four labels into two labels.

```{r}
all_trips <- all_trips %>% 
  mutate(member_casual = recode(member_casual, "Subscriber" = "member", 
                                "Customer" = "casual"))
unique(all_trips$member_casual)
```

-   We can aggregate the data at the ride-level. And for that, we need
    to add some additional columns such as day, month, year using
    started_at column.

```{r}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
colnames(all_trips)
```

-   Now adding a calculated field to calculate the length of the ride as
    this column was removed from the data starting 2020.

```{r}
all_trips$ride_length <- difftime(all_trips$ended_at, all_trips$started_at)
```

```{r Checking structure of the columns}
str(all_trips)
```

Ride length is different datatype so converting it to numeric so that
can perform calculations.

```{r Changing data type to numberic}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

-   There are few rides where trip duration was negative, which includes
    rides where the Divvy took bikes out of circulation for maintenance.
    We want to remove those many rides. So creating new version of data
    frame(v2)

```{r Deleting rides with negative trip duration}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length < 0),]

```

### Descriptive Analysis

Descriptive analysis on ride_length column(in seconds).

```{r Finding Mean, median, max and minimum}
summary(all_trips_v2$ride_length)
```

Descriptive analysis on member_casual Column (Char type)

```{r Compaaring member and casual riders}
aggregate(all_trips_v2$ride_length~all_trips_v2$member_casual, FUN=mean)
aggregate(all_trips_v2$ride_length~all_trips_v2$member_casual, FUN=median)
aggregate(all_trips_v2$ride_length~all_trips_v2$member_casual, FUN=max)
aggregate(all_trips_v2$ride_length~all_trips_v2$member_casual, FUN=min)
```

Average ride time for members vs casual riders by each day

```{r}
aggregate(all_trips_v2$ride_length~all_trips_v2$member_casual+ all_trips_v2$day_of_week, FUN=mean)

```

Days of the week seems out of order so fixing it

```{r Fixing the order and rechecking}
all_trips_v2$day_of_week<- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday","Tuesday","Wednesday","Thursday","Friday","Saturday"))
# rechecking the order
aggregate(all_trips_v2$ride_length~all_trips_v2$member_casual+ all_trips_v2$day_of_week, FUN=mean)

```

Now analyzing riders data based on Type and weekday

```{r}
all_trips_v2 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual,day_of_week)
```

Let's visualize these numbers

```{r}
all_trips_v2 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, day_of_week) %>% 
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```

Let's visualize the number of rides by average duration

```{r}
all_trips_v2 %>% 
  mutate(weekday=wday(started_at,label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual,weekday) %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

Let's visualize by rider type and month

```{r}
all_trips_v2 %>% 
  group_by(member_casual, month) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual,month) %>% 
  ggplot(aes(x = month, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```

Let's visualize the by average duration throught out the year.

```{r}
all_trips_v2 %>% 
  group_by(member_casual, month) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual,month) %>% 
  ggplot(aes(x = month, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

### Exporting summary file for further analysis

```{r}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
write.csv(counts, file = 'avg_ride_length.csv')

```

**Conclusion**

-   Number of rides through out the week is more for members vs causal
-   Less number of rides but more length duration for causal vs members
-   Number of rides are more from Apr-Oct for both member and causal.

**Recommendations**

-   *Can start the rewards program for the membership sign- up. And
    start some campaigns to attract more users to sign up.*

-   As casual riders take longer trips, we can offer cheaper ride when
    member vs causal.
